---
title: "Permutation Feature Importance applied to XGBoost"
author: "José Alberto Arcos Sánchez"
date: "4 de enero de 2018"
output: html_notebook
---

# Abstract

In this document I will calculate the importance of each feature in a complex model (XGBoost) using **Permutation Feature Importance**. For more information about **Permutation Feature Importance**, go and read the great book [Interpretable Machine Learning](https://christophm.github.io/interpretable-ml-book/permutation-feature-importance.html).


<br><br>

# Load the model, the data and the needed packages

```{r}
library(tidyverse)
library(ggplot2)
library(magrittr)

load("../../0-model-to-explain/xgboost-titanic.RData")
```


<br><br>

# Calculate base accuracy

```{r}
predictions_ref <- predict(xgb_model, preprocessed_test_set)
predictions_ref <- as.integer(predictions_ref > 0.5)
acc_ref <- 
  sum(predictions_ref == test_set$survived, na.rm = TRUE) /
  length(predictions_ref)
acc_ref
```



<br><br>

# Permute each feature and calculate accuracy

## Function for continuous variables

```{r}

feature_importance <- function(feature_name, test_data, model,
                               labels, n_permutations, acc_ref){
  
  importances <- numeric(n_permutations)
  feat_index <- which(colnames(test_data) == feature_name)
  
  for(i in 1:n_permutations){
    permutation <- sample(x = 1:nrow(test_data))
    test_data[, feat_index] <- test_data[permutation, feat_index]
    
    predictions <- predict(model, test_data)
    predictions <- as.integer(predictions > 0.5)
    acc <- sum(predictions == labels, na.rm = TRUE) / length(predictions)
    
    importances[i] <- acc_ref - acc
  }
  
  importances
  
}

```

## Function for Pclass

```{r}

feature_importance_pclass <- function(test_data, model, 
                                      classes = test_set$Pclass, 
                                      labels, n_permutations, acc_ref){
  
  importances <- numeric(n_permutations)
  pclass2_index <- which(colnames(test_data) == "Pclass.2")
  pclass3_index <- which(colnames(test_data) == "Pclass.3")
  
  for(i in 1:n_permutations){
    class <- sample(classes) # class permutation
    pclass2_permutation <- as.integer(class == 2)
    pclass3_permutation <- as.integer(class == 3)
    
    test_data[, pclass2_index] <- pclass2_permutation
    test_data[, pclass3_index] <- pclass3_permutation
    
    predictions <- predict(model, test_data)
    predictions <- as.integer(predictions > 0.5)
    acc <- sum(predictions == labels, na.rm = TRUE) / length(predictions)
    
    importances[i] <- acc_ref - acc
  }
  
  importances
  
}

```

## Function for Embarked

```{r}

feature_importance_embarked <- function(test_data, model, 
                                        classes = test_set$Embarked, 
                                        labels, n_permutations, acc_ref){
  
  importances <- numeric(n_permutations)
  embarkedC_index <- which(colnames(test_data) == "Embarked.C")
  embarkedQ_index <- which(colnames(test_data) == "Embarked.Q")
  embarkedS_index <- which(colnames(test_data) == "Embarked.S")
  
  for(i in 1:n_permutations){
    class <- sample(classes) # class permutation
    embarkedC_permutation <- as.integer(class == "C")
    embarkedQ_permutation <- as.integer(class == "Q")
    embarkedS_permutation <- as.integer(class == "S")
    
    test_data[, embarkedC_index] <- embarkedC_permutation
    test_data[, embarkedQ_index] <- embarkedQ_permutation
    test_data[, embarkedS_index] <- embarkedS_permutation
    
    predictions <- predict(model, test_data)
    predictions <- as.integer(predictions > 0.5)
    acc <- sum(predictions == labels, na.rm = TRUE) / length(predictions)
    
    importances[i] <- acc_ref - acc
  }
  
  importances
  
}

```

## Age

```{r}
set.seed(23)
age_importance <- 
  feature_importance(feature_name = "Age",
                     test_data = preprocessed_test_set,
                     labels = test_set$survived,
                     model = xgb_model,
                     n_permutations = 400,
                     acc_ref = acc_ref)

age_importance
```


## Fare

```{r}
set.seed(23)
fare_importance <- 
  feature_importance(feature_name = "Fare",
                     test_data = preprocessed_test_set,
                     labels = test_set$survived,
                     model = xgb_model,
                     n_permutations = 400,
                     acc_ref = acc_ref)

fare_importance
```


## Sibilings

```{r}
set.seed(23)
sibsp_importance <- 
  feature_importance(feature_name = "SibSp",
                     test_data = preprocessed_test_set,
                     labels = test_set$survived,
                     model = xgb_model,
                     n_permutations = 400,
                     acc_ref = acc_ref)

sibsp_importance
```


## Sex

```{r}
set.seed(23)
sex_importance <- 
  feature_importance(feature_name = "male",
                     test_data = preprocessed_test_set,
                     labels = test_set$survived,
                     model = xgb_model,
                     n_permutations = 400,
                     acc_ref = acc_ref)

sex_importance
```


## Parch

```{r}
set.seed(23)
parch_importance <- 
  feature_importance(feature_name = "Parch",
                     test_data = preprocessed_test_set,
                     labels = test_set$survived,
                     model = xgb_model,
                     n_permutations = 400,
                     acc_ref = acc_ref)

parch_importance
```


# Pclass

```{r}
set.seed(23)
pclass_importance <-
  feature_importance_pclass(test_data = preprocessed_test_set,
                            model = xgb_model,
                            n_permutations = 400,
                            labels = test_set$survived,
                            acc_ref = acc_ref)
pclass_importance
```


# Embarked

```{r}
set.seed(23)
embarked_importance <-
  feature_importance_embarked(test_data = preprocessed_test_set,
                              model = xgb_model,
                              n_permutations = 400,
                              labels = test_set$survived,
                              acc_ref = acc_ref)
embarked_importance
```


# Results

```{r}
importance <- data.frame(Pclass = pclass_importance,
                         Parch = parch_importance,
                         Sex = sex_importance,
                         SibSp = sibsp_importance, 
                         Fare = fare_importance, 
                         Age = age_importance,
                         Embarked = embarked_importance)
```

```{r}
importance <- tidyr::gather(data = importance, 
                            key = "feature", 
                            value = "importance")
levels <- importance %>% 
  group_by(feature) %>% 
  summarise(median_importance = median(importance)) %>% 
  arrange(median_importance) %>% .$feature
  
importance$feature <- factor(x = importance$feature, 
                             levels = levels)
```

```{r}
ggplot(data = importance, mapping = aes(x = feature, y = importance)) +
  geom_boxplot(fill = "green") +
  geom_hline(yintercept = 0, color = "red", linetype = "dotted") +
  coord_flip() + 
  theme_minimal()
```